HERA Dish Beam Measurements Memo

### 2015-10-28 DishMeasurements/memo/DishBeamMeasurementsMemo.tex

\title{HERA Dish Beam Measurements Memo, Rev. $\alpha$}

@@

\author{Abraham Neben, Rich Bradley, Jackie Hewitt}
%\affil{MIT Kavli Institute, Massachusetts Institute of Technology, Cambridge, MA, 02139 USA}

\author{Oct 25, 2015}
%\affil{Wasamatta U}

@@

\begin{abstract}
write abstract here
comment on how this expt is still a work in progress
\end{abstract}

@@

%\section{Other title ideas}
%\begin{itemize}
%\item The relationship between noise correlation and monopole sensitivity in radio interferometers
%\item The self noise of sky noise in monopole interferometers
%\item Signal and noise in global signal interferometers
%\end{itemize}

@@

@@

We review the

\begin{figure*}[h]
\includegraphics[width=6.5in]{aerial.png}
\caption{Aerial photograph of the two reference dipoles deployed in Galford Meadow at the National Radio Astronomy Observatory--Green Bank 100\,m south of the HERA dish.}
\label{fig:null3}
\end{figure*}

specify which of the three HERA dishes we are talking about

sky coverage with orbcomm experiment

\subsection{Reference antenna model}
Our power ratio measurement relies upon an accurately modeled reference dipole, and to that end we improve on the model used in \citep{neben15} by XXX

\subsection{Dish models}
\label{sec:dishmodels}

We compare measured beam patterns with a simplistic analytic model and two numerical electromagnetic models. NEED MORE DETAILS ABOUT MODELING
\begin{enumerate}
	\item Airy pattern for a 14\,m aperture
	\item CST simulation (\texttt{hera-cst/HERA\_DISH\_paper\_feed\_cyl36\_140mhz\_Y\_healpix.fits})
	\item A separate simulation run by Dave
\end{enumerate}

\section{Null Experiments}

@@

\begin{figure}[h]
\includegraphics[width=3.5in]{measbeam_raw.png}
\includegraphics[width=3.5in]{measbeam_interp.png}
\includegraphics[width=3.5in]{measbeam_interp_expcutoff.png}
\caption{We construct full sky beams as required for Section \ref{sec:sci} by starting from the nominal focus beam (top left), interpolating over unobserved cells at $\theta\lesssim60^\circ$ and extrapolating to the horizon (top right), and applying a gaussian cutoff starting at $\theta=60^\circ$ with $\sigma=2.5^\circ$ (bottom left). These last two beams span the space of likely horizon responses.}
\label{fig:interpbeams}
\end{figure}

@@

\section{Implications for HERA Science}
\label{sec:sci}

\subsection{Foreground Leakage}
for a power spectrum analysis, note that these tests test how leakage varies as a function of the brightness of emission at different delays, not leakage due to beam frequency dependence (ie, either pattern or overall gain)

\begin{figure*}[h]
\includegraphics[width=5.5in]{LST0deg_14m_42m_EWbaselinesblackman.png}
\includegraphics[width=5.5in]{LST60deg_14m_42m_EWbaselinesblackman.png}
\caption{aeuaoeuaoeu}
\label{fig:delayspec}
\end{figure*}

\begin{figure*}[h]
\includegraphics[width=5.5in]{LST0deg_14m_42m_NSslantbaselinesblackman.png}
\includegraphics[width=5.5in]{LST0deg_14m_42m_SEslantbaselinesblackman.png}
\caption{aeuaoeuaoeu}
\label{fig:delayspec2}
\end{figure*}

\subsection{Foreground Residuals}
the accuracy of beam modeling affects the science performance in additional ways ranging from missubtraction to imperfect foreground subtraction (Neben et al. 2015b submitted). The impact of beam mismodeling on science performance is left as future work.

\section{Discussion}

As expected, the Airy patter has the largest collecting area

the beam might be narrower simply because more dish is illuminated, but not necessarily a good thing, because we want more taper at the edge

comment on how this expt is still a work in progress

>>>>

\title{HERA Dish Beam Measurements Memo}

@@

\author{Abraham Neben, Rich Bradley, Jackie Hewitt, others}
%\affil{MIT Kavli Institute, Massachusetts Institute of Technology, Cambridge, MA, 02139 USA}

\author{Oct 26, 2015}

@@

\begin{abstract}
We deploy the 137\,MHz ORBCOMM beam mapping system of \citet{neben15} at the site of the HERA prototype at NRAO--Green Bank. This technique measures the beam of an antenna-under-test relative to that of a well-modeled reference antenna. We characterize environmental systematics such as reflections and multipath effects by comparing the measured beams of different reference antennas, then measure the beam pattern of the east-most HERA dish as a function of feed height over the dish surface. With the feed at the nominal focus of 4.5\,m over the dish surface, the collecting area is observed to be 68.5\,m$^2$, agreeing with simulations. We also simulate delay spectra on baselines of different lengths and orientations at different LSTs for measured and model beams, and quantify the severity of and uncertainties in the delay space horizon brightening termed the ``pitchfork effect'' by \citet{nithya15}. Future measurements will study the dish beam in the presence of adjacent dishes, quantifying levels of cross-talk and cross-coupling.
\end{abstract}

@@

@@

%must use frequency structure to separate FGs from signal
%a certain amount of freq structure is imprinted onto smooth sources due to the frequency-dependent sampling function of an interferometer
%also by faraday rotation and the bandpass (citations)

%leakage is due to bandpass, frequency flagging, finite bandwidth. can be mitigated by subtraction in the delay spectrum (CLEAN), giving up on wedge modes, or subtraction in image space with foreground modeling, likely will have to give up on wedge modes too unless have incredible calibration and imaging fidelity, as LOFAR is attempting. Large risk of signal loss and calibration-induced transfer of spectral structure from long baselines to short ones

%hera is pursuing two relatively orthogonal power spectrum analyses (imaging and delay transform)

%the other papers in this series quantify the degree of spectral structure induced by the dish bandpass in measurement and modeling, we focus in this section on foreground leakage due to the finite bandwidth as a function of the beam response at different delays in the sky. Nithya discuss this is detail nithya15,nithya15b

%\citet{nithya15,nithya15b} 
%for a power spectrum analysis, note that these tests test how leakage varies as a function of the brightness of emission at different delays, not leakage due to beam frequency dependence (ie, either pattern or overall gain)

\section{ORBCOMM Beam Mapping System Review}

We briefly review the beam mapping system detailed by \citet{neben15}, then discuss application of system to the prototype HERA array at NRAO--Green Bank. The system takes advantage of the 137\,MHz communications satellites operated by ORBCOMM Inc. as bright point sources which, by virtue of their number ($\sim30$), short orbital periods ($\sim90$ minutes), and orbital precession cover 65\% of the visible sky in just a few days. The coverage is limited by the fact that the satellites' orbital inclinations are all less than $45^\circ$.

In contrast to celestial source beam measurements, though, where the flux may be assumed constant over the timescale of the measurement, satellite fluxes can vary rapidly due to varying distance, orientation, and transmission power. To correct for this, we measure the satellite flux in each ground polarization (EW and NS) using a simple, well-modeled reference antenna. Comparison of this measured power with that observed in the Antenna-Under-Test (AUT) gives the AUT beam response in the direction of the satellite. An equivalent interpretation is that the power ratio between the AUT and the reference antenna gives the relative beam response in the satellite direction, and multiplication by the reference antenna model yields the desired AUT response. As discussed in \citet{neben15}, despite the fact that satellite signals are generally polarized, this procedure.

In detail, we measure the dual-polarization RMS powers for each antenna in 512 2\,kHz bands across the 137--138\,MHz band. Each channel power is averaged over $\sim0.2$\,sec. There are 0--3 satellites above the horizon at any given time transmitting on different $\sim15$\,kHz wide sub-bands in 137--138\,MHz. By observing at many different frequencies, we probe the beam response in all these directions simultaneously. We compute the satellite positions using the orbital elements published by Celestrak\footnote{http://www.celestrak.com/NORAD/elements/orbcomm.txt} and the orbital integrator \texttt{predict}\footnote{http://www.qsl.net/kd2bd/predict.html}. However, the satellite frequencies vary occasionally to avoid interference within the constellation. \citet{zheng14} use interferometric phases to identify and exclude times when multiple satellites are in view. As our data acquisition system makes only total power measurements, we instead use an ORBCOMM interface box (typically supplied to commercial users of the network) to sync with passing satellites and record their identifier and transmission frequency.

In this way, beam measurements are built up along satellite tracks over the course of several days of integration, yielding typically 200--300 satellite pass. Each pass is processed separately to identify and exclude times of low signal-to-background when the satellite is low in the sky or in the off state of of a pulsing sequence. At those times, then satellite flux no longer dominates over that of the diffuse Galactic background, and a power measurement no longer probes the response in the satellite direction. The beam measurements are then gridded in horizontal coordinates in HEALPix with a resolution of $1.8^\circ$ (nside=32). As a last quality control step to reject errant beam measurements due to RFI, for instance, we keep only the central 90\% of $\sim50$ measured beam values in each HEALPix cell.

\subsection{Reference Antenna Modeling}
Our power ratio measurement relies upon an accurately modeled reference dipole. We are developing a more comprehensive model of the physical reference antenna and ground screen design, improving on that used in \citep{neben15}. As yet, this improved model yields incorrect results, so more debugging is needed.

\subsection{Dish and Feed Modeling}
\label{sec:dishmodels}

We compare measured beam patterns with a simplistic analytic model and two numerical electromagnetic models. Briefly, the dish is constructed of overlapping mesh rectangles secured to hanging PVC pipes to form an approximately parabolic surface with diameter 14\,m. The feed consists of a dual-polarization dipole antenna with a circular mesh backplane and a cylindrical mesh ``skirt'' hanging down. The feed is suspended over the dish surface by three ropes attached to poles equally spaced around the dish.

\begin{enumerate}
	\item Airy pattern for a 14\,m aperture
	\item CST simulation (\texttt{hera-cst/HERA\_DISH\_paper\_feed\_cyl36\_140mhz\_Y\_healpix.fits})
	\item A separate simulation run by Dave with 68'' diameter feed backplane and 26'' skirt.
\end{enumerate}

@@

A 3-element HERA engineering prototype array is being constructed at the National Radio Astronomy Observatory--Green Bank. We performed the beam measurements presented in this work on the first of these to be constructed, future work will characterize how its in-situ beam in the presence of the other two dishes once they are constructed. Figure \ref{fig:aerial} shows the west-most HERA dish next to two planned dishes in Galford Meadow. The two reference antennas 100\,m are positioned due south with a 100\,m separation, and the Green Bank Telescope $\sim1$\,km northwest of the dish. The data acquisition system is installed in the hut labeled ``Launch point'' on the image.

Note that unlike the full HERA site in the Karoo Desert Radio Astronomy Reserve in South Africa, the Green Bank site has obstructions such as trees and foothills, as well as moist ground with possibly non-uniform properties. All these will contribute to variation of the in-situ reference antenna relative to the electromagnetic model. We will discuss below how we quantify these systematics with ``null experiments'' which compare the beams of the two reference antennas.

\begin{figure*}[h]
\includegraphics[width=6.5in]{aerial.png}
\caption{Aerial photograph of the two reference dipoles deployed in Galford Meadow at the National Radio Astronomy Observatory--Green Bank 100\,m south of the HERA dish.}
\label{fig:aerial}
\end{figure*}

\section{Assessing Systematics with a Null Experiment}

@@

\begin{figure}[h]
\includegraphics[width=3.5in]{measbeam_raw.png}
\includegraphics[width=3.5in]{measbeam_interp.png}
\includegraphics[width=3.5in]{measbeam_interp_expcutoff.png}
\caption{We construct full sky beams as required for Section \ref{sec:foregroundleakage} by starting from the nominal focus beam (top left), interpolating over unobserved cells at $\theta\lesssim60^\circ$ and extrapolating to the horizon (top right), and applying a gaussian cutoff starting at $\theta=60^\circ$ with $\sigma=2.5^\circ$ (bottom left). These last two beams span the space of likely horizon responses. Circles denote $\theta=20^\circ,40^\circ,60^\circ,80^\circ$.}
\label{fig:interpbeams}
\end{figure}

@@

Why does the collecting area increase as the feed is raised over the nominal focus? There are two computing effects here: (1) the reflected radiation is less well focused above the feed, decreasing the response; and (2) the tapered feed sees a larger dish as it is raised, allowing more radiation to reach the dipole as opposed to reflecting off the skirt. These data suggest that the second effect wins. Of course the reason for the skirt is to taper the feed beam in order to mitigate cross-talk and cross-coupling between the dishes. A larger collecting area may not be worth it in exchange for exacerbating these concerns.

\section{Foreground Leakage and Residuals}
\label{sec:foregroundleakage}

We assess the level of foreground leakage into the EOR window due to beam shape, baseline length, and LST. In particular, we built on the study of \citet{nithya15} of the impact of beam shape and horizon response on foreground spectral structure and containment in the wedge.

We simulate delay spectra for several model and measured beams on different baseline lengths for different LSTs and baseline orientations. All these factors affect the level of emission horizon brightening in delay space, the ``pitchfork'' effect, and the level of foreground leakage. We assume a constant beam pattern over that band, with leakage due only to the finite bandwidth and choice of window function. Actual frequency structure imprinted onto the foregrounds by variation of the beam angular pattern or overall gain with frequency (Patra et al.)

We begin by simulating visibilities over a 20\,MHz bandwidth with  400\,kHz resolution centered on 137\,MHz on two different baselines for two model beams (airy, and CST sim) and two measured beams (measured nominal focus beam, and measured nominal focus beam with $60^\circ$ cutoff). We use the short and maximally redundant 14\,m baseline, most useful for delay spectrum analysis, as well as a longer 42\,m baseline, more useful for imaging analysis. From these frequency-sequenced visibilities we calculate the delay spectrum  \citep{perbaselinetechnique} using the Blackman-Harris window following \citet{nithya15}. This window reduces the noise equivalent bandwidth to 10\,MHz, corresponding to $\Delta z\sim0.5$, beyond which evolution of the cosmological signal become significant. Larger bandwidths, though, are useful for foreground estimation and delay-space deconvolution \citep{parsonsandbacker,paper32,paper64}. With this caveat, our simulated delay spectra show the worst case scenario, quantifying the level of foreground isolation without estimation and deconvolution.

Our sky model is comprised of a deep MWA point source survey within 20$^\circ$ of R.A.(J2000) $= 0^\text{h}\,0^\text{m}\,0^\text{s}$ and decl.(J2000) $= -30^\circ\,0'\,0''$ (Carroll et al., in prep.), the shallower but wider MWA commissioning point source survey\citep{MWACS}, the Culgoora catalog\citep{Slee1995}, and the Global Sky Model of Galactic radio emission \citep{gsm}.

Figure \ref{fig:delayspec} shows the delay spectra at $0^\circ$ (top pair) and $60^\circ$ (bottom pair) LST for 14\,m (left pair) and 42\,m (right pair) EW baselines. For reference we plot a 1D model 21\,m power spectrum for $z\sim8$ from \citet{21cmfast}. Consider first the $0^\circ$ LST plots where the Galactic disk has nearly, but not completely set. The effect of horizon response is most visible on the longer 42\,m baseline where the measured beam (red) and CST model (yellow), both show significant brightening at the horizon delays (vertical black lines) due to roughly -30\,dB beam response near the horizon, as discussed by \citet{nithya15,nithya15}. In contrast, the airy model (black) and measured beam with cutoff (blue) show no significant brightening at the horizon, where they appear still to be dominated by the wings of the delay PSF from the emission closer to zero delay.

We also plot (green dashed line) the delay spectrum after subtracting the visibilities computed with the $\theta\sim60^\circ$ cutoff measured beam from those using the full measured beam. This curve may be interpreted as the uncertainty in the ``pitchfork'' horizon brightening due to uncertainty in the beam horizon response, or equivalently as the delay spectrum residual foregrounds after subtracting foregrounds in the well modeled region of the sky. By construction, this subtraction leaves emission from the highest delay regions of the sky (near the horizon on opposite sides of the baseline), but also leaves some low delay emission on opposite sides of a line perpendicular to the baseline. However, this additional low delay emission near the horizon gets drowned out by that from near zenith where the beam is substantially stronger.

On the shorter 14\,m baseline, the horizon brightening is smaller because there is roughly an order of magnitude more emission at zero delay, whose thereby stronger delay wings, which are also relatively wider compared to the narrower foreground wedge, cover up much of the edge brightening. At $60^\circ$ LST, effect of horizon response is much smaller as the Galactic disk is now entirely below the horizon.

The level of leakage depends also on baseline orientation which sets which directions on the sky map to large delays. Figure \ref{fig:delayspec2} shows the delay spectra at $0^\circ$ LST on 14\,m NE (top pair) and SE (bottom pair) baselines. Both show increased edge brightening compared to the 14\,m EW baseline. \citet{nithya15} proposes to identify through modeling the baselines most susceptible to this edge brightening at any given time, and simply exclude them from the power spectrum analysis.

\begin{figure*}[h]
\includegraphics[width=6.7in]{LST0deg_14m_42m_EWbaselines_dish1_blackman.png}
\includegraphics[width=6.7in]{LST60deg_14m_42m_EWbaselines_dish1_blackman.png}
\caption{Simulated delay spectra for the GSM and point sources at $0^\circ$ (top pair) and $60^\circ$ (bottom pair) LST for EW baselines of length 14\,m (left pair) and 42\,m (right pair). The beams are are assumed constant over frequency.}
\label{fig:delayspec}
\end{figure*}

\begin{figure*}[h]
\includegraphics[width=6.7in]{LST0deg_14m_42m_NEslantbaselines_dish1_blackman.png}
\includegraphics[width=6.7in]{LST0deg_14m_42m_SEslantbaselines_dish1_blackman.png}
\caption{Simulated delay spectra for the GSM and point sources at $0^\circ$ LST for NE (top pair) and SE (bottom pair) baselines of length 14\,m (left pair) and 42\,m (right pair). The beams are are assumed constant over frequency.}
\label{fig:delayspec2}
\end{figure*}

%\section{Discussion}

%As expected, the Airy patter has the largest collecting area

%the beam might be narrower simply because more dish is illuminated, but not necessarily a good thing, because we want more taper at the edge

%comment on how this expt is still a work in progress
